On the propriety of restricted Boltzmann machines

### 2016-11-14 writing/draft.Rmd
more vardeman edits

### Joint distribution
Let $\boldsymbol x = (h_1, \dots, h_\nh, v_1,\dots,v_\nv)$ represent the states of the visible and hidden nodes in an RBM for some integers $\nv, \nh \ge 1$. Each single binary random variable, visible or hidden, will take its values in a common coding set $\mathcal{C}$, where we allow one of two possibilities for the coding set, $\mathcal{C}=\{0,1\}$ or $\mathcal{C}=\{-1,1\}$, with "$1$" always indicating the "high" value of the variable. While $\mathcal{C}=\{0,1\}$ may be a natural or naive starting point, we argue later that the coding $\mathcal{C}=\{-1,1\}$ induces more interpretable model properties for the RBM. A standard parametric form for probabilities corresponding to a potential vector of states, $X = (H_1, \dots, H_\nh, V_1,\dots,V_\nv)$, for the nodes is

@@

We next explore and numerically explain the relationship between values of $\boldsymbol \theta$ and the three measures of model impropriety, near-degeneracy, instability, and uninterpretability, for RBM models of varying size.

@@

For this small model, we are able to investigate the symptoms of model impropriety, beginning with near-degeneracy. To this end, recall that one characterization requires the convex hull of statistics 
$$\mathcal{T} = \{\boldsymbol t(\boldsymbol x): \boldsymbol x = (v_1, h_1) \in \{0,1\}^2\} \equiv \{(v_1, h_1, v_1 h_1): v_1, h_1 \in \{0,1\}\}$$ 
appearing in the RBM probabilities for this model. As this set is in three dimensions, we are able to explicitly illustrate the shape of boundary of the convex hull of $\mathcal{T}$ and explore the behavior of the mean parameterization of the parameters. Figure \ref{fig:toyhull} shows the convex hull of our "statistic space," $\mathcal{T} \subset \{0,1\}^3$, for this toy problem from two perspectives (enclosed by the unit cube, representing the convex hull of $\{0,1\}^3$ space). In this small model, note that the convex hull of $\mathcal{T}$ does not fill the unrestricted hull of $\{0,1\}^3$ because of the relationship between the elements of $\mathcal{T} = \{(v_1, h_1, v_1 h_1 : v_1, h_1 \in \{0, 1\} \}$ (i.1. $v_1 h_1 = 1$ only if $v_1 = h_1 = 1$).

@@

    ```{r, results='asis', echo=FALSE, include=TRUE, message=FALSE}
    cat(sapply(1:nrow(m1_bin$c_hull$hull), function(x) paste0("\\draw", paste(paste0("(", apply(m1_bin$possible_t[cbind(m1_bin$c_hull$hull, m1_bin$c_hull$hull[,1])[x,],], 1, paste, collapse=","), ")"), collapse=" -- "), "; \n")))
    ```
    \end{tikzpicture}
  }
  \end{minipage}
  \caption{The convex hull of the "statistic space"" in three dimensions for the toy RBM with one visible and one hidden node. The convex hull of $\mathcal{T} = \{\boldsymbol t(\boldsymbol x) : \boldsymbol x \in \mathcal{C}^{\nv + \nh}\}$ does not fill the unit cube because of the relationship between the elements of $\mathcal{T}$.}
\label{fig:toyhull}
\end{figure}

Additionally, we can compute the mean vector for $\boldsymbol t(\boldsymbol x)$ as a function of the model parameters as 
\begin{align}
\boldsymbol \mu(\boldsymbol \theta) &= \text{E}_{\boldsymbol \theta}\left[ \boldsymbol t(\boldsymbol X) \right] =  \sum\limits_{\boldsymbol x = (v_1, h_1) \in \mathcal{X} = \{0,1\}^2} \left\{ t(x) \frac{\exp\left( \theta_{11} h_1 v_1 + \theta_{h1} h_1 + \theta_{v1} v_1\right)}{\gamma(\boldsymbol \theta)} \right\}
= \left[
\begin{matrix}
\frac{\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)}{\gamma(\boldsymbol \theta)} \\
\frac{\exp\left(\theta_{h_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)}{\gamma(\boldsymbol \theta)} \\
\frac{\exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)}{\gamma(\boldsymbol \theta)} \\
\end{matrix} \right]
\end{align}
where $\gamma(\boldsymbol \theta) = \sum\limits_{h_1 = 0}^1\sum\limits_{v_1 = 0}^1 \exp(\theta_{11} h_1 v_1 + \theta_{h_1}h_1 + \theta_{v_1}v_1)$. The three parametric coordinate functions of $\boldsymbol \mu(\boldsymbol \theta)$ can be visualized in Figure \ref{fig:degen_toy}. (Contour plots for three coordinate functions are shown in columns for various values of $\theta_{11}$.) In examining these, we see that as values of coordinates of $\boldsymbol \theta$ grow larger in magnitude, at least one mean function for the entries of $\boldsymbol t(\boldsymbol x)$ is approaching a value 0 or 1, forcing $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol x)$ to be near to the boundary of the convex hull of $\mathcal{T}$, as a sign of model near-degeneracy. Thus, for a very small example we can see the relationship between values of $\boldsymbol \theta$ and model degeneracy.

@@

Finally to consider the effect of $\boldsymbol \theta$ on potential model uninterpretability, we can look at the difference between model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and expectations given independence, E$\left[\boldsymbol X | \boldsymbol \theta^*\right ]$ for the tiny toy RBM model where $\boldsymbol X = (V_1, H_1, V_1 H_1)$. This is given by

\begin{align*}
\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] -   \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right] 
&= \left[
\begin{matrix}
\frac{\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)}{\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)} \\
\frac{\exp\left(\theta_{h_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)}{\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)} 
\end{matrix} \right] - \left[
\begin{matrix}
\frac{\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{v_1} + \theta_{h_1}\right)}{\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(\theta_{v_1} + \theta_{h_1}\right)} \\
\frac{\exp\left(\theta_{h_1}\right) + \exp\left(+ \theta_{v_1} + \theta_{h_1}\right)}{\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(+ \theta_{v_1} + \theta_{h_1}\right)} 
\end{matrix} \right] \\
&= \left[
\begin{matrix}
\frac{\left|\exp\left(\theta_{11} + \theta_{v_1} + 2\theta_{h_1}\right) - \exp\left( \theta_{v_1} + 2\theta_{h_1}\right) \right| }{\left(\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)\right)\left(\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(+ \theta_{v_1} + \theta_{h_1}\right)\right)} \\
\frac{\left|\exp\left(\theta_{11} + 2\theta_{v_1} + \theta_{h_1}\right) - \exp\left( 2\theta_{v_1} + \theta_{h_1}\right) \right| }{\left(\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)\right)\left(\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(+ \theta_{v_1} + \theta_{h_1}\right)\right)}
\end{matrix}\right].
\end{align*}

@@

```{r uninterp, fig.cap="\\label{fig:uninterp}The difference between coordinates of model expectations, E$\\left[\\boldsymbol X | \\boldsymbol \\theta\\right]$, and expectations given independence, E$\\left[\\boldsymbol X | \\boldsymbol \\theta^* \\right ]$ for a RBM with one visible and one hidden node.", fig.height = 4, fig.align='center', message=FALSE, cache=TRUE, warning=FALSE}
expand.grid(theta_h1 = seq(-5, 5, length.out = 21),
            theta_v1 = seq(-5, 5, length.out = 21),
            theta_11 = seq(-5, 5, length.out = 5)) %>%
  data.frame() %>%
  group_by(theta_v1, theta_h1, theta_11) %>%
  do(data.frame(gamma = sum(exp(possibles %*% t(data.matrix(.)))),
                mu_v1_dep = possibles[,"v1"] %*% exp(possibles %*% t(data.matrix(.))),
                mu_h1_dep = possibles[,"h1"] %*% exp(possibles %*% t(data.matrix(.))))) %>%
  mutate_each(funs(./gamma), starts_with("mu")) %>%
  select(-gamma) -> exp_dep

@@

### Data coding to mitigate apparent degeneracy
Multiple encodings of the binary variables are possible. For example, we could allow hiddens $(H_1, \dots, H_\nh) \in \{0,1\}^\nh$ and visibles $(V_1, \dots, V_\nv) \in \{0,1\}^\nv$, as in the previous sections or we could instead encode the state of the variables as $\{-1,1\}^\nh$ and $\{-1,1\}^\nv$. This will result in variables $\boldsymbol t(\boldsymbol X)$ from (\ref{eqn:t}) satisfying $\boldsymbol t(\boldsymbol x) \in \{0,1\}^{\nh + \nv + \nh*\nv}$ or $\boldsymbol t(\boldsymbol x) \in \{-1,1\}^{\nh + \nv + \nh*\nv}$ depending on how we encode "on" and "off" states in the nodes. To explore the effect of this encoding on the potential for apparent near-degeneracy of an RBM, we can consider the ratio of the volume of the hypercube with corners at elements of $\mathcal{T}$ to the volume of the convex hull of $\mathcal{T}$ under both possible encodings (i.e., compare convex hull of $\mathcal{T}$ to either $[0,1]^3$ or $[-1,1]^3$ under the respective encodings).

For the small two node example, the $0/1$ encoding loses `r round((1-m1_bin$c_hull$vol)*100,2)`% of the volume of $[0,1]^3$ and the $-1/1$ encoding loses `r round((1-m1_neg$c_hull$vol/(2^3))*100,2)`% of the volume of $[-1,1]^3$. While there is clearly a one-to-one mapping of models for these two possible codings, this notion of lost volume can be helpful to conceptualize how difficult it will be for the mean vector $\boldsymbol \mu(\boldsymbol \theta)$ to avoid the boundary of $\mathcal{T}$, and thus avoid apparent near-degeneracy. In fact, if we look at the ratio of volume within the convex hull defined by $\mathcal{T}$ and the corresponding hypercube, we can see a relationship emerge as the number of nodes (and thus parameters) increases. In Figure \ref{fig:volume_plot}, it is evident that as the number of parameters increases, this ratio is decreasing at an increasing rate, meaning it gets more and more difficult to avoid the boundary of the convex hull and thus appearance of near degeneracy. Additionally, it appears that the $-1/1$ encoding suffers slightly less from this problem. This suggests that if intuition about models is to be formed in terms of a data encoding that encourages of "non-pathological" means $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}(\boldsymbol t(\boldsymbol x))$ for $\boldsymbol t(\boldsymbol x)$, then the $-1/1$ encoding is preferable to $0/1$ coding.

@@

In addition to the argument that the $-1/1$ data encoding mitigates some perceived prevalence of near-degeneracy it also has the benefit of providing a guaranteed-to-be non-degenerate model at $\boldsymbol \theta = \boldsymbol 0 \in \mathbb{R}^{\nh + \nv + \nh*\nv}$, where the zero vector that serves as the natural center of the parameter space and induces the simplest possible model properties for the RBM (i.e., at $\boldsymbol \theta = \boldsymbol 0$, all variables are independent and visible variables are independent and uniformly distributed on $\{-1,1\}^\nv$). The proof of this is included in appendix \ref{center-of-universe}. Hence while from some computing perspectives $0/1$ coding might seem most natural, the $-1/1$ coding is far more convenient and interpretable from the point of view of statistical modeling, where it makes parameters simply interpreted in terms of symmetrically defined main effects and interactions. In light of all of these matters we will henceforth employ the $-1/1$ coding.

@@

To explore the effects of the RBM parameter vector $\boldsymbol \theta$ as it relates to near-degeneracy, instability, and uninterpretability, we consider models of small size. For $\nh, \nv \in \{1, \dots, 4\}$, we sample `r dim(res$samp[[1]])[1]` values of $\boldsymbol \theta$ with various magnitudes (details to follow). For each set of parameters we then calculate the three metrics of model impropriety introduced in Section \ref{degeneracy-instability-and-uninterpretability-oh-my}, $\boldsymbol \mu(\boldsymbol \theta)$, $\text{ELPR}(\boldsymbol \theta)/\nv$, and the coordinates of $\left|\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] -   \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right] \right|$. In the case of near-degeneracy, we can go further and classify each model as near-degenerate or "viable" based on the distance of $\boldsymbol \mu(\boldsymbol \theta)$ from the boundary of the convex hull of $\mathcal{T}$ and look at the fraction of models that are "near-degenerate," meaning they are within a small distance to the boundary of the convex hull. We define a "small" through a rough estimation to the volume of the hull for each model size. We pick $\epsilon_0 = 0.05$ for $\nh=\nv=1$ and then, for every other $\nh$ and $\nv$, set $m=\nh+\nv+\nv*\nh$ and pick $\epsilon$ so that $1-(1-2\epsilon_0)^3 = 1 - (1-2\epsilon)^m$. In this way, we roughly scale the volume of the small distance to the boundary of the convex hull to be equivalent in all dimensions.

In this numerical experiment, we split $\boldsymbol \theta$ into $\boldsymbol \theta_{main}$ and $\boldsymbol \theta_{interaction}$, in reference to which variables in the probability function the parameters correspond to (whether they multiply a $v_i$ or a $h_j$ or they multiply a $v_i h_j$), and allow the two types of terms to have varying average magnitudes, $||\boldsymbol \theta_{main} || /(\nh+\nv)$ and $||\boldsymbol \theta_{interaction} || /(\nh*\nv)$. These average magnitudes vary on a grid between `r min(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V))` and `r max(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V))` with `r length(unique(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V)))` breaks, yielding `r length(unique(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V)))^2` grid points. (By looking at the average magnitudes, we are able to later consider the potential benefit of shrinking each parameter value $\theta_i$ towards zero in a Bayesian fitting technique.) At each point in the grid, `r dim(res$samp[[1]])[1]` vectors are uniformly sampled on a sphere with radius corresponding to the first coordinate in the grid ($\boldsymbol \theta_{main}$) and `r dim(res$samp[[1]])[1]` vectors are uniformly sampled on a sphere with radius corresponding to the second coordinate in the grid ($\boldsymbol \theta_{main}$) via sums of squared and scaled iid Normal$(0, 1)$ variables. These vectors are then paired to create `r dim(res$samp[[1]])[1]` values of $\boldsymbol \theta$ with magnitudes at each point in the grid.

The results of this numerical study are summarized in Figures \ref{fig:degen_plots}, \ref{fig:instab_plots}, and \ref{fig:uninterp_plots}. From these three figures, it is clear that all three measures of model impropriety show higher values for larger magnitudes of the parameter vectors. Additionally, since there are $\nh*\nv$ interaction terms in $\boldsymbol \theta$ versus only $\nh + \nv$ main effect terms, for large models there are many more interaction parameters than main effects in the model. And so, severely limiting the magnitude of the individual interactions may well help prevent model impropriety.

>>>>

### Joint distribution
Let $\boldsymbol x = (h_1, \dots, h_\nh, v_1,\dots,v_\nv)$ represent the states of the visible and hidden nodes in a RBM for some integers $\nv, \nh \ge 1$. Each single binary random variable, visible or hidden, will take its values in a common coding set $\mathcal{C}$, where we allow one of two possibilities for the coding set, $\mathcal{C}=\{0,1\}$ or $\mathcal{C}=\{-1,1\}$, with "$1$" always indicating the "high" value of the variable. While $\mathcal{C}=\{0,1\}$ may be a natural or naive starting point, we argue later that the coding $\mathcal{C}=\{-1,1\}$ induces more interpretable model properties for the RBM. A standard parametric form for probabilities corresponding to a potential vector of states, $X = (H_1, \dots, H_\nh, V_1,\dots,V_\nv)$, for the nodes is

@@

We next explore and numerically explain the relationship between values of $\boldsymbol \theta$ and the three measures of model impropriety, near-degeneracy, instability, and uninterpretability, for RBM models of varying sizes.

@@

For this small model, we are able to investigate the symptoms of model impropriety, beginning with near-degeneracy. To this end, recall that one characterization requires consideration of the convex hull of possible values of statistics $\boldsymbol t(\boldsymbol x)$,
$$\mathcal{T} = \{\boldsymbol t(\boldsymbol x): \boldsymbol x = (v_1, h_1) \in \{0,1\}^2\} \equiv \{(v_1, h_1, v_1 h_1): v_1, h_1 \in \{0,1\}\}$$ 
appearing in the RBM probabilities for this model. As this set is in three dimensions, we are able to explicitly illustrate the shape of boundary of the convex hull of $\mathcal{T}$ and explore the behavior of the mean vector $\text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol x)$ as a function of the parameter vector $\boldsymbol \theta$. Figure \ref{fig:toyhull} shows the convex hull of our "statistic space," $\mathcal{T} \subset \{0,1\}^3$, for this toy problem from two perspectives (enclosed by the unit cube, representing the convex hull of $\{0,1\}^3$ space). In this small model, note that the convex hull of $\mathcal{T}$ does not fill the unrestricted hull of $\{0,1\}^3$ because of the relationship between the elements of $\mathcal{T} = \{(v_1, h_1, v_1 h_1 : v_1, h_1 \in \{0, 1\} \}$ (i.1. $v_1 h_1 = 1$ only if $v_1 = h_1 = 1$).

@@

    ```{r, results='asis', echo=FALSE, include=TRUE, message=FALSE}
    cat(sapply(1:nrow(m1_bin$c_hull$hull), function(x) paste0("\\draw", paste(paste0("(", apply(m1_bin$possible_t[cbind(m1_bin$c_hull$hull, m1_bin$c_hull$hull[,1])[x,],], 1, paste, collapse=","), ")"), collapse=" -- "), "; \n")))
    ```
    \end{tikzpicture}
  }
  \end{minipage}
  \caption{The convex hull of the "statistic space" in three dimensions for the toy RBM with one visible and one hidden node. The convex hull of $\mathcal{T} = \{\boldsymbol t(\boldsymbol x) : \boldsymbol x \in \mathcal{C}^{\nv + \nh}\}$ does not fill the unit cube because of the relationship between the elements of $\mathcal{T}$.}
\label{fig:toyhull}
\end{figure}

We can compute the mean vector for $\boldsymbol t(\boldsymbol x)$ as a function of the model parameters as 
\begin{align}
\boldsymbol \mu(\boldsymbol \theta) &= \text{E}_{\boldsymbol \theta}\left[ \boldsymbol t(\boldsymbol X) \right] =  \sum\limits_{\boldsymbol x = (v_1, h_1) \in \mathcal{X} = \{0,1\}^2} \left\{ t(x) \frac{\exp\left( \theta_{11} h_1 v_1 + \theta_{h1} h_1 + \theta_{v1} v_1\right)}{\gamma(\boldsymbol \theta)} \right\}
= \left[
\begin{matrix}
\frac{\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)}{\gamma(\boldsymbol \theta)} \\
\frac{\exp\left(\theta_{h_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)}{\gamma(\boldsymbol \theta)} \\
\frac{\exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)}{\gamma(\boldsymbol \theta)} \\
\end{matrix} \right]
\end{align}
where $\gamma(\boldsymbol \theta) = \sum\limits_{h_1 = 0}^1\sum\limits_{v_1 = 0}^1 \exp(\theta_{11} h_1 v_1 + \theta_{h_1}h_1 + \theta_{v_1}v_1)$. The three parametric coordinate functions of $\boldsymbol \mu(\boldsymbol \theta)$ can be represented as in Figure \ref{fig:degen_toy}. (Contour plots for three coordinate functions are shown in columns for various values of $\theta_{11}$.) In examining these, we see that as values of coordinates of $\boldsymbol \theta$ grow larger in magnitude, at least one mean function for the entries of $\boldsymbol t(\boldsymbol x)$ is approaching a value 0 or 1, forcing $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol x)$ to be near to the boundary of the convex hull of $\mathcal{T}$, as a sign of model near-degeneracy. Thus, for a very small example we can see the relationship between values of $\boldsymbol \theta$ and model degeneracy.

@@

Finally to consider the effect of $\boldsymbol \theta$ on potential model uninterpretability, we can look at the absolute difference between model expectations, E$\left[\boldsymbol X | \boldsymbol \theta\right]$, and expectations given independence, E$\left[\boldsymbol X | \boldsymbol \theta^*\right ]$ for the tiny toy RBM model where $\boldsymbol X = (V_1, H_1, V_1 H_1)$. This difference is given by

\begin{align*}
\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] -   \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right] 
&= \left[
\begin{matrix}
\frac{\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)}{\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)} \\
\frac{\exp\left(\theta_{h_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)}{\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)} 
\end{matrix} \right] - \left[
\begin{matrix}
\frac{\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{v_1} + \theta_{h_1}\right)}{\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(\theta_{v_1} + \theta_{h_1}\right)} \\
\frac{\exp\left(\theta_{h_1}\right) + \exp\left(+ \theta_{v_1} + \theta_{h_1}\right)}{\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(+ \theta_{v_1} + \theta_{h_1}\right)} 
\end{matrix} \right] \\
&= \left[
\begin{matrix}
\frac{\exp\left(\theta_{11} + \theta_{v_1} + 2\theta_{h_1}\right) - \exp\left( \theta_{v_1} + 2\theta_{h_1}\right) }{\left(\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)\right)\left(\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(+ \theta_{v_1} + \theta_{h_1}\right)\right)} \\
\frac{\exp\left(\theta_{11} + 2\theta_{v_1} + \theta_{h_1}\right) - \exp\left( 2\theta_{v_1} + \theta_{h_1}\right)  }{\left(\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(\theta_{11} + \theta_{v_1} + \theta_{h_1}\right)\right)\left(\exp\left(\theta_{v_1}\right) + \exp\left(\theta_{h_1}\right) + \exp\left(+ \theta_{v_1} + \theta_{h_1}\right)\right)}
\end{matrix}\right].
\end{align*}

@@

```{r uninterp, fig.cap="\\label{fig:uninterp}The absolute difference between coordinates of model expectations, E$\\left[\\boldsymbol X | \\boldsymbol \\theta\\right]$, and expectations given independence, E$\\left[\\boldsymbol X | \\boldsymbol \\theta^* \\right ]$ for a RBM with one visible and one hidden node.", fig.height = 4, fig.align='center', message=FALSE, cache=TRUE, warning=FALSE}
expand.grid(theta_h1 = seq(-5, 5, length.out = 21),
            theta_v1 = seq(-5, 5, length.out = 21),
            theta_11 = seq(-5, 5, length.out = 5)) %>%
  data.frame() %>%
  group_by(theta_v1, theta_h1, theta_11) %>%
  do(data.frame(gamma = sum(exp(possibles %*% t(data.matrix(.)))),
                mu_v1_dep = possibles[,"v1"] %*% exp(possibles %*% t(data.matrix(.))),
                mu_h1_dep = possibles[,"h1"] %*% exp(possibles %*% t(data.matrix(.))))) %>%
  mutate_each(funs(./gamma), starts_with("mu")) %>%
  select(-gamma) -> exp_dep

@@

### Data coding to mitigate apparent degeneracy
Multiple encodings of the binary variables are possible. For example, we could allow hiddens $(H_1, \dots, H_\nh) \in \{0,1\}^\nh$ and visibles $(V_1, \dots, V_\nv) \in \{0,1\}^\nv$, as in the previous sections or we could instead encode the state of the variables as $\{-1,1\}^\nh$ and $\{-1,1\}^\nv$. This will result in variables $\boldsymbol t(\boldsymbol X)$ from (\ref{eqn:t}) satisfying $\boldsymbol t(\boldsymbol x) \in \{0,1\}^{\nh + \nv + \nh*\nv}$ or $\boldsymbol t(\boldsymbol x) \in \{-1,1\}^{\nh + \nv + \nh*\nv}$ depending on how we encode "on" and "off" states in the nodes. To explore the effect of this encoding on the potential for apparent near-degeneracy of a RBM, we can consider the ratio of the volume of the hypercube with corners at elements of $\mathcal{T}$ to the volume of the convex hull of $\mathcal{T}$ under both possible encodings (i.e., compare convex hull of $\mathcal{T}$ to either $[0,1]^3$ or $[-1,1]^3$ under the respective encodings).

For the small two node example, the $0/1$ encoding loses `r round((1-m1_bin$c_hull$vol)*100,2)`% of the volume of $[0,1]^3$ and the $-1/1$ encoding loses `r round((1-m1_neg$c_hull$vol/(2^3))*100,2)`% of the volume of $[-1,1]^3$. While there is clearly a one-to-one mapping between models for these two possible codings, this notion of lost volume can be helpful to geometrically conceptualize how difficult it will be for the mean vector $\boldsymbol \mu(\boldsymbol \theta)$ to avoid the boundary of $\mathcal{T}$, and thus avoid apparent near-degeneracy. In fact, if we look at the ratio of volume within the convex hull defined by $\mathcal{T}$ and the corresponding hypercube, we can see a relationship emerge as the number of nodes (and thus parameters) increases. In Figure \ref{fig:volume_plot}, it is evident that as the number of parameters increases, this ratio is decreasing at an increasing rate, meaning it gets more and more difficult to avoid the boundary of the convex hull and thus appearance of near degeneracy. Additionally, it appears that the $-1/1$ encoding suffers slightly less from this problem. This suggests that if intuition about models is to be formed in terms of a data encoding that encourages of "non-pathological" means $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}(\boldsymbol t(\boldsymbol x))$ for $\boldsymbol t(\boldsymbol x)$, then the $-1/1$ encoding is preferable to $0/1$ coding.

@@

In addition to the argument that the $-1/1$ data encoding mitigates some perceived prevalence of near-degeneracy it also has the benefit of providing a guaranteed-to-be non-degenerate model at $\boldsymbol \theta = \boldsymbol 0 \in \mathbb{R}^{\nh + \nv + \nh*\nv}$, where the zero vector that serves as the natural center of the parameter space and induces the simplest possible model properties for the RBM (i.e., at $\boldsymbol \theta = \boldsymbol 0$, all variables are independent and visible variables are independent and uniformly distributed on $\{-1,1\}^\nv$). The proof of this and further exploration o the equivalence of the $\boldsymbol \theta$ parameterization of the RBM model class and parameterization by $\boldsymbol \mu(\boldsymbol \theta)$ is in appendix \ref{center-of-universe}. Hence while from some computing perspectives $0/1$ coding might seem most natural, the $-1/1$ coding is far more convenient and interpretable from the point of view of statistical modeling, where it makes parameters simply interpreted in terms of symmetrically defined main effects and interactions. In light of all of these matters we will henceforth employ the $-1/1$ coding.

@@

To explore the impact of RBM parameter vector $\boldsymbol \theta$ magnitude on near-degeneracy, instability, and uninterpretability, we consider models of small size. For $\nh, \nv \in \{1, \dots, 4\}$, we sample `r dim(res$samp[[1]])[1]` values of $\boldsymbol \theta$ with various magnitudes (details to follow). For each set of parameters we then calculate metrics of model impropriety introduced in Section \ref{degeneracy-instability-and-uninterpretability-oh-my} based on $\boldsymbol \mu(\boldsymbol \theta)$, $\text{ELPR}(\boldsymbol \theta)/\nv$, and the coordinates of $\left|\text{E}\left[\boldsymbol X | \boldsymbol \theta\right] -   \text{E}\left[\boldsymbol X | \boldsymbol \theta^* \right] \right|$. In the case of near-degeneracy, we classify each model as near-degenerate or "viable" based on the distance of $\boldsymbol \mu(\boldsymbol \theta)$ from the boundary of the convex hull of $\mathcal{T}$ and look at the fraction of models that are "near-degenerate," meaning they are within a small distance of the boundary of the convex hull. We define a "small" through a rough estimation of the volume of the hull for each model size. We pick $\epsilon_0 = 0.05$ for $\nh=\nv=1$ and then, for every other $\nh$ and $\nv$, set $m=\nh+\nv+\nv*\nh$ and pick $\epsilon$ so that $1-(1-2\epsilon_0)^3 = 1 - (1-2\epsilon)^m$. In this way, we roughly scale the volume of the "small distance" to the boundary of the convex hull to be equivalent across dimensions.

In our numerical experiment, we split $\boldsymbol \theta$ into $\boldsymbol \theta_{main}$ and $\boldsymbol \theta_{interaction}$, in reference to which variables in the probability function the parameters correspond (whether they multiply a $v_i$ or a $h_j$ or they multiply a $v_i h_j$), and allow the two types of terms to have varying average magnitudes, $||\boldsymbol \theta_{main} || /(\nh+\nv)$ and $||\boldsymbol \theta_{interaction} || /(\nh*\nv)$. These average magnitudes vary on a grid between `r min(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V))` and `r max(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V))` with `r length(unique(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V)))` breaks, yielding `r length(unique(plot_dat_grid$r1/(plot_dat_grid$H + plot_dat_grid$V)))^2` grid points. (By looking at the average magnitudes, we are able to later consider the potential benefit of shrinking each parameter value $\theta_i$ towards zero in a Bayesian fitting technique.) At each point in the grid, `r dim(res$samp[[1]])[1]` vectors are uniformly sampled on a sphere with radius corresponding to the first coordinate in the grid ($\boldsymbol \theta_{main}$) and `r dim(res$samp[[1]])[1]` vectors are uniformly sampled on a sphere with radius corresponding to the second coordinate in the grid ($\boldsymbol \theta_{main}$) via sums of squared and scaled iid Normal$(0, 1)$ variables. These vectors are then paired to create `r dim(res$samp[[1]])[1]` values of $\boldsymbol \theta$ with magnitudes at each point in the grid.

The results of this numerical study are summarized in Figures \ref{fig:degen_plots}, \ref{fig:instab_plots}, and \ref{fig:uninterp_plots}. From these three figures, it is clear that all three measures of model impropriety show higher values for larger magnitudes of the parameter vectors. Additionally, since there are $\nh*\nv$ interaction terms in $\boldsymbol \theta$ versus only $\nh + \nv$ main effect terms, for large models there are many more interaction parameters than main effects in the models. And so, severely limiting the magnitude of the individual interactions may well help prevent model impropriety.

```{r degen_plots, fig.cap="\\label{fig:degen_plots}Results from the numerical experiment, here looking at the fraction of models that were near-degenerate for each combination of magnitude of $\\boldsymbol \\theta$ and model size. Black lines show the contour levels for fraction of near-degeneracy, while the thick black line shows the level where the fraction of near-degenerate models is .05.", fig.height=4}
three_ways %>%
  mutate(r1 = round(r1/(H + V), 8), r2 = round(r2/(H * V), 8)) %>% 
  ggplot() +
  geom_tile(aes(x = r1, y = r2, fill = frac_degen)) +
  geom_contour(aes(x = r1, y = r2, z = frac_degen), colour = "black", bins = 8) +
  geom_contour(aes(x = r1, y = r2, z = frac_degen), colour = "black", breaks = .05, size = 1.5) +
  geom_abline(aes(intercept = 0, slope = 1), alpha = .5, lty = 2) +
  scale_fill_gradient("Fraction near-degenerate", low = "yellow", high = "red") +
  facet_grid(Visibles~Hiddens, labeller = label_parsed) +
  xlab(expression(paste(paste("||", theta[main], "||", "/(", n[h]," + ", n[v], ")")))) +
  ylab(expression(paste(paste("||", theta[interaction], "||", "/(", n[h], "*", n[v], ")")))) +
  theme(aspect.ratio = 1) -> p.degen
